---
title: "Null models"
author: "Mackenzie Johnson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Relationship between $\lambda$ and $n_{eff}$ under a null model

This is the RMarkdown document for the null models used to produce Figure 5. One null model has amino acid data generated under an exponential distribution, while the other was simulated from a normal distribution.

Load packages
```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(cowplot)
library(broom)
library(ggtext)
theme_set(theme_cowplot())

```

Generate data under an exponential and normal distribution
```{r, message=FALSE}
###### exponential ######
# generate data into tidy_align df
# correct method: multinomial with expo prob function
# p(k)=ce^(-lambda*k)
# rmultinom(1, 500, exp(-1*(0:19)))
# 1 site with 500 aa counts
# range of lambda to get range of eff n

# generate 110 values of lambda evenly spaced between 0.15 and 1
lambda_exp <- seq(0.10, 1, length.out = 110)

# function for multinomial sampling with exponential probability function
sample_null_exp <- function(lambda){
  
  # 1 site with 500 sequences
  null_site_exp <- t(rmultinom(1, 500, exp(-lambda*(0:19))))
  
}

# get count data for each lambda value
null_exp <- lapply(lambda_exp, sample_null_exp)

# convert list to dataframe
null_exp <- data.frame(
  matrix(
    unlist(null_exp), 
    nrow=length(null_exp), 
    byrow=T
  )
)

# recreate all_count df (site, aa, count)
null_exp %>% 
  mutate(site = seq(1:nrow(null_exp))) %>% 
  pivot_longer(
    -site, 
    names_to = "aa", 
    values_to = "count"
  ) -> all_count_exp

###### normal dist ######
# correct method: multinomial with normal prob function
# p(k)=ce^(-(k-10)^2/(sigma^2))
# range of sigman, range of neff
# 1 site with 500 aa counts

# generate 110 values of sigma evenly spaced between 0.5 and 15
sigma_nrm <- seq(0.5, 15, length.out = 110)

# function for multinomial sampling with normal probability function
sample_null_nrm <- function(sigma){
  
  # 1 site with 500 sequences
  null_site_nrm <- t(rmultinom(1, 500, exp(-1*((0:19)-10)^2/(sigma^2))))
  
}

# get count data for each sigma value
null_nrm <- lapply(sigma_nrm, sample_null_nrm)

# convert list to dataframe
null_nrm <- data.frame(
  matrix(
    unlist(null_nrm), 
    nrow=length(null_nrm), 
    byrow=T
  )
)

# recreate all_count df (site, aa, count)
null_nrm %>% 
  mutate(site = seq(1:nrow(null_nrm))) %>% 
  pivot_longer(
    -site, 
    names_to = "aa", 
    values_to = "count"
  ) -> all_count_nrm


```


Estimate distribution
```{r}
# USING ONLY OBSERVED AA
# order aa by relative frequency
all_count_exp %>%
  arrange(all_count_exp$site, desc(count)) -> ordered_count_exp
all_count_nrm %>%
  arrange(all_count_nrm$site, desc(count)) -> ordered_count_nrm

# make the categorical variable (aa) numerical (k) by numbering aa's 1-20 in order of freq
ordered_count_exp %>%
  group_by(site) %>%
  mutate(k = as.numeric(0:19)) -> ordered_count_exp
ordered_count_nrm %>%
  group_by(site) %>%
  mutate(k = as.numeric(0:19)) -> ordered_count_nrm

# remove aa with zero values
ordered_count_exp %>%
  filter(count > 0) -> observed_aa_exp
ordered_count_nrm %>%
  filter(count > 0) -> observed_aa_nrm

# count relative to most frequent - rescale to 1 for highly conserved sites
observed_aa_exp %>%
  group_by(site) %>%
  mutate(rel_count = count/max(count)) -> observed_aa_exp
observed_aa_nrm %>%
  group_by(site) %>%
  mutate(rel_count = count/max(count)) -> observed_aa_nrm

# log-transformed data
observed_aa_exp %>%
  mutate(ln_count = log(rel_count)) -> observed_aa_exp 
observed_aa_nrm %>%
  mutate(ln_count = log(rel_count)) -> observed_aa_nrm 

# fit to a linear function, using map and BROOM 
# SET INTERCEPT TO 0
observed_aa_exp %>% 
  nest(data = c(aa, count, k, rel_count, ln_count)) %>%
  mutate(
    fit = map(data, ~ lm(ln_count ~ 0 + k, data = .)),
    slope = map_dbl(fit, ~ (.)$coefficients[1])
  ) %>% 
  select(-data, -fit) -> observed_fits_exp
observed_aa_nrm %>% 
  nest(data = c(aa, count, k, rel_count, ln_count)) %>%
  mutate(
    fit = map(data, ~ lm(ln_count ~ 0 + k, data = .)),
    slope = map_dbl(fit, ~ (.)$coefficients[1])
  ) %>% 
  select(-data, -fit) -> observed_fits_nrm

ordered_count_exp %>% 
  left_join(observed_fits_exp, by = "site") %>%
  mutate(est_dist = exp(k*slope)) -> ordered_count_exp
ordered_count_nrm %>% 
  left_join(observed_fits_nrm, by = "site") %>%
  mutate(est_dist = exp(k*slope)) -> ordered_count_nrm

# rescale to compare w raw count
ordered_count_exp %>%
  group_by(site) %>%
  mutate(est_rel = est_dist/sum(est_dist)) %>%
  mutate(est_count = sum(count)*est_rel) -> ordered_count_exp
ordered_count_nrm %>%
  group_by(site) %>%
  mutate(est_rel = est_dist/sum(est_dist)) %>%
  mutate(est_count = sum(count)*est_rel) -> ordered_count_nrm


```


Chi-squared: null vs. estimated distribution
```{r}
# exponential
ordered_count_exp %>% 
  nest(data = c(aa, count, k, slope, est_dist, est_rel, est_count)) %>%
  mutate(
    chisq = map(data, ~ sum(((.$count - .$est_count)^2)/.$est_count)),
    p_value = map_dbl(chisq, ~ pchisq(., 18, lower.tail=FALSE))
  ) %>% 
  select(-data) %>%
  mutate(p.adjusted = p.adjust(p_value, method= "fdr")) -> chisq_results_exp

chisq_results_exp %>%
  mutate(result = "pass") -> chisq_results_exp

chisq_results_exp$result[chisq_results_exp$p.adjusted < 0.05] <- "fail"

chisq_results_exp %>% filter(result == "fail") %>% nrow()/nrow(chisq_results_exp)

# normal
ordered_count_nrm %>% 
  nest(data = c(aa, count, k, slope, est_dist, est_rel, est_count)) %>%
  mutate(
    chisq = map(data, ~ sum(((.$count - .$est_count)^2)/.$est_count)),
    p_value = map_dbl(chisq, ~ pchisq(., 18, lower.tail=FALSE))
  ) %>% 
  select(-data) %>%
  mutate(p.adjusted = p.adjust(p_value, method= "fdr")) -> chisq_results_nrm

chisq_results_nrm %>%
  mutate(result = "pass") -> chisq_results_nrm

chisq_results_nrm$result[chisq_results_nrm$p.adjusted < 0.05] <- "fail"

chisq_results_nrm %>% filter(result == "fail") %>% nrow()/nrow(chisq_results_nrm)


```


Effective number of amino acids
```{r}
# eff aa for actual distribution
observed_aa_exp %>%
  group_by(site) %>%
  mutate(frequency = (count)/sum(count)) -> site_aa_freq_exp
observed_aa_nrm %>%
  group_by(site) %>%
  mutate(frequency = (count)/sum(count)) -> site_aa_freq_nrm

site_aa_freq_exp %>% 
  mutate(flnf = frequency*log(frequency)) %>%
  group_by(site) %>% 
  summarize(
    entropy = -sum(flnf),
    eff_aa = exp(entropy),
    n = length(unique(aa))
  ) -> eff_aa_all_exp
site_aa_freq_nrm %>% 
  mutate(flnf = frequency*log(frequency)) %>%
  group_by(site) %>% 
  summarize(
    entropy = -sum(flnf),
    eff_aa = exp(entropy),
    n = length(unique(aa))
  ) -> eff_aa_all_nrm

# eff aa for estimated distribution
ordered_count_exp %>% 
  group_by(site) %>%
  mutate(freq = (est_count)/sum(est_count)) -> site_aa_freq_2_exp
site_aa_freq_2_exp %>% 
  mutate(flnf = freq*log(freq)) %>%
  group_by(site) %>%
  summarize(
    entr = -sum(flnf),
    eff_aa_est = exp(entr)
  ) -> eff_aa_all_est_exp
ordered_count_nrm %>% 
  group_by(site) %>%
  mutate(freq = (est_count)/sum(est_count)) -> site_aa_freq_2_nrm
site_aa_freq_2_nrm %>% 
  mutate(flnf = freq*log(freq)) %>%
  group_by(site) %>%
  summarize(
    entr = -sum(flnf),
    eff_aa_est = exp(entr)
  ) -> eff_aa_all_est_nrm

left_join(eff_aa_all_exp, eff_aa_all_est_exp) -> compare_eff_aa_exp
left_join(eff_aa_all_nrm, eff_aa_all_est_nrm) -> compare_eff_aa_nrm

```


Relationship between $\lambda$ and $n_{eff}$
```{r}
observed_fits_exp %>%
  left_join(eff_aa_all_exp) %>% 
  select(site, slope, eff_aa)-> gam_v_effaa_exp
observed_fits_nrm %>%
  left_join(eff_aa_all_nrm) %>% 
  select(site, slope, eff_aa)-> gam_v_effaa_nrm

gam_v_effaa_exp %>%
  left_join(chisq_results_exp) %>% 
  select(site, slope, eff_aa, result) -> gam_v_effaa_exp
gam_v_effaa_nrm %>%
  left_join(chisq_results_nrm) %>% 
  select(site, slope, eff_aa, result) -> gam_v_effaa_nrm

# theoretical prediction
neff <- function(lambda_inv) {
  lambda <- 1/lambda_inv
  pi <- lambda*exp(-lambda*(0:19))
  C <- sum(pi)
  pi <- pi/C
  exp(-sum(pi*log(pi)))
}
#neff2(1/3)

df_theory <- data.frame(
  lambda_inv = (0:-400)/10, 
  ne = vapply((0:-400)/10, neff, numeric(1))
)

gam_v_effaa_exp %>% 
  mutate(null = "Positive control") -> gam_v_effaa_exp
gam_v_effaa_nrm %>% 
  mutate(null = "Negative control") -> gam_v_effaa_nrm

full_join(
  gam_v_effaa_exp, 
  gam_v_effaa_nrm, 
  by = c("site", "slope", "eff_aa", "result", "null")
) -> gam_v_effaa

```


Figure 5
```{r}
# reorder panels
gam_v_effaa$null <- factor(
  gam_v_effaa$null, 
  levels = c("Positive control", "Negative control")
)

ggplot(gam_v_effaa, 
       aes(x = eff_aa, 
           y = -1/slope)) + 
  geom_point(aes(color = result)) +
  geom_line(
    data = df_theory, 
    aes(ne, -1*lambda_inv), 
    inherit.aes = FALSE, 
    size = 1
  ) +
  facet_wrap(vars(null)) +
  labs(
    x = "Effective number of amino acids *n*<sub>eff</sub>", 
    y = "Fitting parameter &lambda;<sup>-1</sup>",
    color = "Result"
  ) +
  scale_color_manual(
    values = c("#E69F00", "#0072B2")
  ) +
  scale_y_continuous(limits = c(-0.5, 31), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0, 20)) +
  theme_cowplot(12, rel_small = 1) +
  panel_border() +
  background_grid() +
  theme(
    legend.position = "right",
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    strip.background = element_rect(fill = "gray90")
  ) -> fig5

fig5

save_plot("figure5.png", fig5, ncol = 1, nrow = 1, base_height = 5,
          base_asp = 1.618, base_width = NULL)


```



