---
title: "Null models"
author: "Mackenzie Johnson"
date: "10/31/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Relationship between $\lambda$ and $n_{eff}$ under a null model

This is the RMarkdown document for the null models used to produce Figure 5. One null model has amino acid data generated under an exponential distribution, while the other was simulated from a normal distribution.

Load packages
```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(cowplot)
library(broom)
library(ggtext)
theme_set(theme_cowplot())

```

Generate data under an exponential and normal distribution
```{r, message=FALSE}
###### exponential ######
# generate data into tidy_align df
# correct method: multinomial with expo prob function
# p(k)=ce^(-lambda*k)
# rmultinom(10, 500, exp(-1*(0:19)))
# range of lambda, range of eff n
# 10 sites with 500 aa counts
null_exp0 <- data.frame(t(rmultinom(10, 500, exp(-1*(0:19))))) 
null_exp1 <- data.frame(t(rmultinom(10, 500, exp(-0.83*(0:19)))))
null_exp2 <- data.frame(t(rmultinom(10, 500, exp(-0.75*(0:19)))))
null_exp3 <- data.frame(t(rmultinom(10, 500, exp(-0.63*(0:19)))))
null_exp4 <- data.frame(t(rmultinom(10, 500, exp(-0.5*(0:19)))))
null_exp5 <- data.frame(t(rmultinom(10, 500, exp(-0.42*(0:19)))))
null_exp6 <- data.frame(t(rmultinom(10, 500, exp(-0.35*(0:19)))))
null_exp7 <- data.frame(t(rmultinom(10, 500, exp(-0.3*(0:19)))))
null_exp8 <- data.frame(t(rmultinom(10, 500, exp(-0.25*(0:19)))))
null_exp9 <- data.frame(t(rmultinom(10, 500, exp(-0.2*(0:19)))))
null_exp10 <- data.frame(t(rmultinom(10, 500, exp(-0.15*(0:19)))))

full_join(null_exp0, null_exp1) %>% full_join(., null_exp2) %>% 
  full_join(., null_exp3) %>% full_join(., null_exp4) %>% 
  full_join(., null_exp5) %>% full_join(., null_exp6) %>%
  full_join(., null_exp7) %>% full_join(., null_exp8) %>% 
  full_join(., null_exp9) %>% full_join(., null_exp10) -> null_exp

# recreate all_count df (site, aa, count)
null_exp %>% 
  mutate(site = seq(1:nrow(null_exp))) %>% 
  pivot_longer(-site, names_to = "aa", values_to = "count") -> all_count_exp

###### normal dist ######
null_nrm0 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.7)))) 
null_nrm1 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.65))))
null_nrm2 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.60))))
null_nrm3 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.55))))
null_nrm4 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.50))))
null_nrm5 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.45))))
null_nrm6 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.40))))
null_nrm7 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.30))))
null_nrm8 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.20))))
null_nrm9 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.10))))
null_nrm10 <- data.frame(t(rmultinom(10, 500, rnorm(20, mean = 1, sd = 0.05))))

full_join(null_nrm0, null_nrm1) %>% full_join(., null_nrm2) %>% 
  full_join(., null_nrm3) %>% full_join(., null_nrm4) %>% 
  full_join(., null_nrm5) %>% full_join(., null_nrm6) %>% 
  full_join(., null_nrm7) %>% full_join(., null_nrm8) %>% 
  full_join(., null_nrm9) %>% full_join(., null_nrm10) -> null_nrm

null_nrm %>% 
  mutate(site = seq(1:nrow(null_nrm))) %>% 
  pivot_longer(-site, names_to = "aa", values_to = "count") -> all_count_nrm


```


Estimate distribution
```{r}
# USING ONLY OBSERVED AA
# order aa by relative frequency
all_count_exp %>%
  arrange(all_count_exp$site, desc(count)) -> ordered_count_exp
all_count_nrm %>%
  arrange(all_count_nrm$site, desc(count)) -> ordered_count_nrm

# make the categorical variable (aa) numerical (k) by numbering aa's 1-20 in order of freq
ordered_count_exp %>%
  group_by(site) %>%
  mutate(k = as.numeric(1:20)) -> ordered_count_exp
ordered_count_nrm %>%
  group_by(site) %>%
  mutate(k = as.numeric(1:20)) -> ordered_count_nrm

# remove aa with zero values
ordered_count_exp %>%
  filter(count > 0) -> observed_aa_exp
ordered_count_nrm %>%
  filter(count > 0) -> observed_aa_nrm

# count relative to most frequent - rescale to 1 for highly conserved sites
observed_aa_exp %>%
  group_by(site) %>%
  mutate(rel_count = count/max(count)) -> observed_aa_exp
observed_aa_nrm %>%
  group_by(site) %>%
  mutate(rel_count = count/max(count)) -> observed_aa_nrm

# log-transformed data
observed_aa_exp %>%
  mutate(ln_count = log(rel_count)) -> observed_aa_exp 
observed_aa_nrm %>%
  mutate(ln_count = log(rel_count)) -> observed_aa_nrm 

# fit to a linear function, using map and BROOM 
# SET INTERCEPT TO 0
observed_aa_exp %>% 
  nest(data = c(aa, count, k, rel_count, ln_count)) %>%
  mutate(fit = map(data, ~ lm(ln_count ~ 0 + k, data = .)),
         #intercept = map_dbl(fit, ~ (.)$coefficients[1]),
         slope = map_dbl(fit, ~ (.)$coefficients[1])) %>% 
  select(-data, -fit) -> observed_fits_exp
observed_aa_nrm %>% 
  nest(data = c(aa, count, k, rel_count, ln_count)) %>%
  mutate(fit = map(data, ~ lm(ln_count ~ 0 + k, data = .)),
         #intercept = map_dbl(fit, ~ (.)$coefficients[1]),
         slope = map_dbl(fit, ~ (.)$coefficients[1])) %>% 
  select(-data, -fit) -> observed_fits_nrm

ordered_count_exp %>% 
  left_join(observed_fits_exp, by = "site") %>%
  mutate(est_dist = exp(k*slope)) -> ordered_count_exp
ordered_count_nrm %>% 
  left_join(observed_fits_nrm, by = "site") %>%
  mutate(est_dist = exp(k*slope)) -> ordered_count_nrm

# rescale to compare w raw count
ordered_count_exp %>%
  group_by(site) %>%
  mutate(est_rel = est_dist/sum(est_dist)) %>%
  mutate(est_count = sum(count)*est_rel) -> ordered_count_exp
ordered_count_nrm %>%
  group_by(site) %>%
  mutate(est_rel = est_dist/sum(est_dist)) %>%
  mutate(est_count = sum(count)*est_rel) -> ordered_count_nrm


```


Chi-squared: null vs. estimated distribution
```{r}
# exponential
ordered_count_exp %>% 
  nest(data = c(aa, count, k, slope, est_dist, est_rel, est_count)) %>%
  mutate(chisq = map(data, ~ sum(((.$count - .$est_count)^2)/.$est_count)),
         p_value = map_dbl(chisq, ~ pchisq(., 18, lower.tail=FALSE))) %>% 
  select(-data) %>%
  mutate(p.adjusted = p.adjust(p_value, method= "fdr")) -> chisq_results_exp

chisq_results_exp %>%
  mutate(result = "pass") -> chisq_results_exp

chisq_results_exp$result[chisq_results_exp$p.adjusted < 0.05] <- "fail"

chisq_results_exp %>% filter(result == "fail") %>% nrow()/nrow(chisq_results_exp)

# normal
ordered_count_nrm %>% 
  nest(data = c(aa, count, k, slope, est_dist, est_rel, est_count)) %>%
  mutate(chisq = map(data, ~ sum(((.$count - .$est_count)^2)/.$est_count)),
         p_value = map_dbl(chisq, ~ pchisq(., 18, lower.tail=FALSE))) %>% 
  select(-data) %>%
  mutate(p.adjusted = p.adjust(p_value, method= "fdr")) -> chisq_results_nrm

chisq_results_nrm %>%
  mutate(result = "pass") -> chisq_results_nrm

chisq_results_nrm$result[chisq_results_nrm$p.adjusted < 0.05] <- "fail"

chisq_results_nrm %>% filter(result == "fail") %>% nrow()/nrow(chisq_results_nrm)


```


Effective number of amino acids
```{r}
# eff aa for actual distribution
observed_aa_exp %>%
  group_by(site) %>%
  mutate(frequency = (count)/sum(count)) -> site_aa_freq_exp
observed_aa_nrm %>%
  group_by(site) %>%
  mutate(frequency = (count)/sum(count)) -> site_aa_freq_nrm

site_aa_freq_exp %>% 
  mutate(flnf = frequency*log(frequency)) %>%
  group_by(site) %>% 
  summarize(entropy = -sum(flnf),
            eff_aa = exp(entropy),
            n = length(unique(aa))) -> eff_aa_all_exp
site_aa_freq_nrm %>% 
  mutate(flnf = frequency*log(frequency)) %>%
  group_by(site) %>% 
  summarize(entropy = -sum(flnf),
            eff_aa = exp(entropy),
            n = length(unique(aa))) -> eff_aa_all_nrm

# eff aa for estimated distribution
ordered_count_exp %>% 
  group_by(site) %>%
  mutate(freq = (est_count)/sum(est_count)) -> site_aa_freq_2_exp
site_aa_freq_2_exp %>% 
  mutate(flnf = freq*log(freq)) %>%
  group_by(site) %>%
  summarize(entr = -sum(flnf),
            eff_aa_est = exp(entr)) -> eff_aa_all_est_exp
ordered_count_nrm %>% 
  group_by(site) %>%
  mutate(freq = (est_count)/sum(est_count)) -> site_aa_freq_2_nrm
site_aa_freq_2_nrm %>% 
  mutate(flnf = freq*log(freq)) %>%
  group_by(site) %>%
  summarize(entr = -sum(flnf),
            eff_aa_est = exp(entr)) -> eff_aa_all_est_nrm

left_join(eff_aa_all_exp, eff_aa_all_est_exp) -> compare_eff_aa_exp
left_join(eff_aa_all_nrm, eff_aa_all_est_nrm) -> compare_eff_aa_nrm

# ggplot(compare_eff_aa_nrm, aes(x=eff_aa, y=eff_aa_est)) + 
#   geom_point() +
#   geom_abline(slope = 1, intercept = 0) + 
#   #xlim(0,16) + 
#   #ylim(0,16) + 
#   labs(title = "Null model linear fit", x = "eff_aa actual", y = "eff_aa fit")

```


Relationship between $\lambda$ and $n_{eff}$
```{r}
observed_fits_exp %>%
  left_join(eff_aa_all_exp) %>% 
  select(site, slope, eff_aa)-> gam_v_effaa_exp
observed_fits_nrm %>%
  left_join(eff_aa_all_nrm) %>% 
  select(site, slope, eff_aa)-> gam_v_effaa_nrm

gam_v_effaa_exp %>%
  left_join(chisq_results_exp) %>% 
  select(site, slope, eff_aa, result) -> gam_v_effaa_exp
gam_v_effaa_nrm %>%
  left_join(chisq_results_nrm) %>% 
  select(site, slope, eff_aa, result) -> gam_v_effaa_nrm

# theoretical prediction
neff <- function(lambda_inv) {
  lambda <- 1/lambda_inv
  pi <- lambda*exp(-lambda*(0:19))
  C <- sum(pi)
  pi <- pi/C
  exp(-sum(pi*log(pi)))
}
#neff2(1/3)

df_theory <- data.frame(lambda_inv = (0:-100)/10, ne = vapply((0:-100)/10, 
                                                              neff, numeric(1)))
df_theory2 <- data.frame(lambda_inv = (0:-400)/10, ne = vapply((0:-400)/10, 
                                                              neff, numeric(1)))

gam_v_effaa_exp %>% 
  mutate(null = "Exponential") -> gam_v_effaa_exp
gam_v_effaa_nrm %>% 
  mutate(null = "Normal") -> gam_v_effaa_nrm

full_join(gam_v_effaa_exp, gam_v_effaa_nrm, by = c("site", "slope", "eff_aa",
                                                   "result", "null")) -> gam_v_effaa

```


Figure 5
```{r}

ggplot(gam_v_effaa, 
       aes(x = eff_aa, 
           y = 1/slope)) + 
  geom_point(aes(color = result)) +
  geom_line(data = df_theory2, 
            aes(ne, lambda_inv), 
            inherit.aes = FALSE, 
            size = 1) +
  facet_wrap(vars(null)) +
  labs(x = "Effective number of amino acids *n*<sub>eff</sub>", 
       y = "Fitting parameter &lambda;<sup>-1</sup>",
       color = "Result") +
  #scale_y_continuous(limits = c(-15, 2), expand = c(0, 0)) +
  theme_cowplot(12, rel_small = 1) +
  panel_border() +
  background_grid() +
  theme(legend.position = "right",
        axis.title.x = element_markdown(),
        axis.title.y = element_markdown(),
        strip.background = element_rect(fill = "gray90")) -> fig5

fig5

save_plot("figure5.png", fig5, ncol = 1, nrow = 1, base_height = 5,
          base_asp = 1.618, base_width = NULL)


```



