---
title: "Simulated protein alignment"
author: "Mackenzie Johnson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of distribution fitting with a single simulated protein alignment

RMarkdown document that shows the process for a single simulated protein alignment, produces Figures 1 and 2.

```{r, message=FALSE}
# fit linear model to normalized, log-transformed data
# set intercept to 0, only fitting 1 parameter (slope)

# load packages
library(tidyverse)
library(stringr)
library(cowplot)
library(broom)
library(ggtext)
theme_set(theme_cowplot())
```

Read in and reformat data
```{r, message=FALSE}
# read in data
alignment <-read_csv("data/simulated/results_1B4T_A_evolved_split.csv", 
                     col_types = cols(.default = "c")) 

# tidy data used for null
alignment %>% 
  mutate(sequence = as.numeric(1:nrow(alignment)) ) %>% 
  gather(key = "site", value = "aa", 1:153) -> tidy_align  

tidy_align$site = as.numeric(tidy_align$site) 
```

Actual distribution
```{r, message=FALSE}
# determine the number of each AA at each site
tidy_align %>% 
  group_by(site, aa) %>% 
  summarize(count = n()) -> site_aa_count

empty_counts <- tibble(
  aa = c('A', 'C', 'D', 'E', 'F', 'G', 'H', 'I','K', 'L', 
         'M', 'N', 'P', 'Q', 'R', 'S', 'T', 'V', 'W', 'Y'),
  count = rep(0, 20)
)

# 1. First write a function that can take a data frame with amino acids at one site, and return a data frame with counts, including zeros.
count_by_site <- function(site_df) {
  site_df %>% 
    group_by(aa) %>% 
    summarize(count = n()) -> counts
  full_join(counts, anti_join(empty_counts, counts, by = 'aa'))
}

# 2. Then call that function for all sites and combine the resulting data frames.
tidy_align %>% 
  nest(data = -site) %>%
  mutate(counts = map(data, count_by_site)) %>%
  dplyr::select(-data) %>%   
  unnest(cols = c(counts)) -> all_count


```

Estimated distribution
```{r, message=FALSE}
# USING ONLY OBSERVED AA
# order aa by relative frequency
all_count %>%
  arrange(all_count$site, desc(count)) -> ordered_count

# make the categorical variable (aa) numerical (k) by numbering aa's 1-20 in order of freq
ordered_count %>%
  group_by(site) %>%
  mutate(k = as.numeric(0:19)) -> ordered_count
  #mutate(k = as.numeric(1:20)) -> ordered_count

# remove aa with zero values
ordered_count %>%
  filter(count > 0) -> observed_aa

# count relative to most frequent - rescale to 1 for highly conserved sites
observed_aa %>%
  group_by(site) %>%
  mutate(rel_count = count/max(count)) -> observed_aa

# log-transformed data
observed_aa %>%
  mutate(ln_count = log(rel_count)) -> observed_aa 

# fit to a linear function, using map and BROOM 
# SET INTERCEPT TO 0
observed_aa %>% 
  nest(data = -site) %>%
  mutate(fit = map(data, ~ lm(ln_count ~ 0 + k, data = .)),
         #intercept = map_dbl(fit, ~ (.)$coefficients[1]),
         slope = map_dbl(fit, ~ (.)$coefficients[1])) %>% 
  select(-data, -fit) -> observed_fits

# data frame with slope from linear fit to log transformed data - slope = gamma, intercept = 0
#head(observed_fits)

# visualize the distribution (at a given site) with linear fits
# NOTE: sites that are highly conserved for a single aa still look exponential in log scale (see sites 3 and 9 as example) 3, 6, 33
ordered_count %>%
  group_by(site) %>%
  mutate(rel_count = count/max(count)) -> ordered_count
ordered_count %>%
  mutate(ln_count = log(rel_count)) -> ordered_count

ordered_count %>% 
  left_join(observed_fits, by = "site") %>%
  mutate(est_dist = slope*exp(k*slope)) -> ordered_count

# rescale to compare w raw count
ordered_count %>%
  group_by(site) %>%
  mutate(est_rel = est_dist/sum(est_dist)) %>%
  mutate(est_count = sum(count)*est_rel) -> ordered_count
```


Chi-squared test to compare distributions
```{r, message=FALSE}
# use raw count
ordered_count %>% nest(data = -site) %>%
  ungroup() %>%
  mutate(chisq = map(data, ~ sum(((.$count - .$est_count)^2)/.$est_count)),
         p_value = map_dbl(chisq, ~ pchisq(., 18, lower.tail=FALSE))) %>% 
  select(-data) %>%
  mutate(p.adjusted = p.adjust(p_value, method= "fdr")) -> chisq_results

chisq_results %>%
  mutate(result = "pass") -> chisq_results

# REMOVE conserved sites with a single aa
chisq_results %>% 
  filter(p_value != "NaN") -> chisq_results

chisq_results$result[chisq_results$p_value < 0.05] <- "fail"

chisq_results %>% filter(result == "fail") %>% nrow()/nrow(chisq_results)
chisq_results$chisq <- as.numeric(chisq_results$chisq)

#test manually for a few sites
#df1 <- filter(ordered_count, site == 1)
#chisq1 <- sum((df1$count - df1$est_count)^2/df1$est_count)
#pchisq(chisq1, 18, lower.tail = FALSE)

```


Figure 1
```{r, message=FALSE}
ordered_count %>%
  filter(site == 35) %>%
  ggplot(aes(x = k)) +
  geom_bar(aes(y = count), fill = "grey", stat='identity') +
  geom_point(aes(y = est_count, color = "violetred")) +
  geom_line(aes(y = est_count, color = "violetred"), size = 1) +
  scale_color_manual(
    name = "distribution",
    values = c("grey" = "grey", "violetred" = "violetred"),
    labels = c("actual", "estimated")
  )  +
  scale_x_continuous(
    breaks = c(0, 5, 10, 15), 
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = c(0, 25, 50, 75, 100), 
    expand = c(0, 0),
    limits = c(0, 108)
  ) +
  labs(x = "Amino acids ranked, *k*", y = "Count") +
  theme(
    legend.position = "none",
    axis.title.x = element_markdown()
  ) -> fig1

fig1
save_plot("figure1.pdf", fig1, ncol = 1, nrow = 1, base_height = 3.71,
          base_asp = 1.618, base_width = NULL)

```

Effective number of amino acids
```{r, message=FALSE}
# site_aa_count is actual dist. DOES NOT CONTAIN AA AT ZERO

# eff aa for actual distribution
observed_aa %>%
  group_by(site) %>%
  mutate(frequency = (count)/sum(count)) -> site_aa_freq

site_aa_freq %>% 
  mutate(flnf = frequency*log(frequency)) %>%
  group_by(site) %>% 
  summarize(entropy = -sum(flnf),
            eff_aa = exp(entropy),
            n = length(unique(aa))) -> eff_aa_all

# eff aa for estimated distribution
ordered_count %>% 
  group_by(site) %>%
  mutate(freq = (est_count)/sum(est_count)) -> site_aa_freq_2
site_aa_freq_2 %>% 
  mutate(flnf = freq*log(freq)) %>%
  group_by(site) %>%
  summarize(entr = -sum(flnf),
            eff_aa_est = exp(entr)) -> eff_aa_all_est

left_join(eff_aa_all, eff_aa_all_est) -> compare_eff_aa

```

Relationship between $\lambda$ and effective number of amino acids
```{r, message=FALSE}
observed_fits %>%
  left_join(eff_aa_all) %>% 
  select(site, slope, eff_aa)-> gam_v_effaa

gam_v_effaa %>%
  left_join(chisq_results) %>% 
  select(site, slope, eff_aa, result) -> gam_v_effaa

neff2 <- function(lambda_inv) {
  lambda <- 1/lambda_inv
  pi <- lambda*exp(-lambda*(0:19))
  #pi <- lambda*exp(-lambda*(1:20))
  C <- sum(pi)
  pi <- pi/C
  exp(-sum(pi*log(pi)))
}

df_theory2 <- data.frame(lambda_inv = (0:-100)/10, 
                         ne = vapply((0:-100)/10, neff2, numeric(1)))
```


Figure 2
```{r, message=FALSE}
gam_v_effaa %>% 
  group_by(result) %>% 
  count()

gam_v_effaa %>% 
  filter(!is.na(result)) -> gam_v_effaa

ggplot(gam_v_effaa, 
       aes(x = eff_aa, y = -1/slope)) + 
  geom_point(aes(color = result)) +
  geom_line(
    data = df_theory2, 
    aes(ne, -1*lambda_inv), 
    inherit.aes = FALSE, 
    size = 1
  ) +
  geom_vline(
    xintercept = 1, 
    linetype = 2
  ) +
  labs(
    x = "Effective number of amino acids *n*<sub>eff</sub>", 
    y = "Fitting parameter &lambda;<sup>-1</sup>",
    color = "Result"
  ) +
  theme(
    #legend.position = "none",
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  ) +
  scale_x_continuous(
    breaks = c(0,5,10,15),
    limits = c(0,17),
    expand = c(0,0)
  ) +
  scale_y_continuous(
    limits = c(-0.1,10),
    expand = c(0,0)
  ) +
  scale_color_manual(
    values = c("#E69F00", "#0072B2")
  ) +
  coord_cartesian(clip = "off") -> fig2

fig2
save_plot("figure2.png", fig2, ncol = 1, nrow = 1, base_height = 3.71,
          base_asp = 1.618, base_width = NULL)

```

